Functors, CPS and monads, or how to generate efficient code from abstract designs

### 2005-03-31 paper/metamonads.tex
some re-writing

\newcommand{\oleg}[1]{{\it [Oleg says: #1]}}

@@

\section{CPS and the problem of generating names}

In this paper, we are concerned with building and composing
generators. We start with the code expressing the algorithm (in
OCaml), and, adding MetaOCaml annotations, turn that code into a
\emph{code expression} that will generate the code when runs. We then
modify that code expression to add various parameters and aspects, to
yield the complete generator.

@@

To solve the problem of writing a generator for a |let| expression 
when the generated body is not yet known, we use CPS.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
And now, a better -- and shorter -- write-up.

@@

The |do|-notation of Haskell (cire report?) shows that it is possible
to write essentially CPS code in the conventional-looking style. The
|do| notation is the notation for writing monadic code (cite Moggi?).
The benefit of monadic notation is not that it can represent CPS (cite
Filinski), but it helps in composability by offering to add different
layers of effects (state, exception, non-determinism, etc) to the
basic monad (cite Hudak, Peyton-Jones).

In the present paper, we use the monad that is the CPS monad with
another layer of state. We will discuss the need for the state in
Section XXX. The type of the monad is:
\begin{code}
type ('v,'s,'w) monad = 's -> ('s -> 'v -> 'w) -> 'w
\end{code}

The monad is just the function of two arguments: the state (of type
|'s|) and the continuation. The continuation receives the current
state, the value (of type |v|) and yields the answer of the type |w|.
The monad is polymorphic over all these three type parameters. Except
a few primitime monad operations below, all other code treat the monad
as an abstract data type.

Monads represent computations that can have ``effects''. In our case,
we can have ``state effects'' (update of state) and ``control effect''
(which occurs whenever the computation does something else but
invoking its continuation once). As we will see, ``control effects''
in our case manifest themselves as ``code generation effects''.

\begin{code}
let ret a = fun s k -> k s a
let bind a f = fun s k -> a s (fun s' b -> f b s' k)
\end{code}

The two monad operations, |ret| and |bind| are conventional: |ret a|
returns a computation that takes the current state and the
continuation and immediately invokes the continuation passing it the
unmodified state and the value. Obviously, the computation |ret a| has
no control effects. The operation |bind a f| combines a
monad |a| of the type |(s,v,w) monad| and the function |f| of the
type |v -> (s,v1,w) monad| into a bigger |(s,v1,w) monad|. The latter
will execute computation |a|, pass it result to |f| and execute the
latter's computation. The state facet of our monad is handled with two
operations:

\begin{code}
let fetch s k = k s s
let store v s k = k (v::s) ()
\end{code}

|fetch| is the computation that yields the current state and |store v|
is the computation that updates the current state and yields |()|. We
model the state as a list (which is actually a list of polymorphic
variants, as we shall see later).

\begin{code}
let retN a = fun s k -> .<let t = .~a in .~(k s .<t>.)>.
\end{code}

@@

Finally, we should be able to run our monad -- that is, perform the
computation of teh monad and get the result -- which in our case is
the code expression. We run the monad by passing it initial state (the
empty list) and the initial continuation |let k0 s v = v|.

We can now re-write our |param_code2| example of the previous section
as

\begin{code}
let param_code3 plus one =
  let gen x y = bind (retN (plus y one)) (fun ce -> 
                ret (plus ce (plus x ce)))
  and run m = m [] k0 
  in .<fun x y -> .~(run (gen .<x>. .<y>.))>.
\end{code}

That does not seem like the improvement. But here we can introduce the
do notation, which is patterned after the do-namtion of Haskell.
\oleg{Explain our mdo briefly? Refer to the message on the
  Ocaml-list?. Refer to camlp4 and note the implementation}

\begin{code}
let param_code4 plus one =
  let gen x y = mdo { ce <-- retN (plus y one);
                      ret (plus ce (plus x ce)) }
  and run m = m [] k0 
  in .<fun x y -> .~(run (gen .<x>. .<y>.))>.
\end{code}

@@

@@

\end{document}

>>>>

\newcommand{\omitnow}[1]{}
\newcommand{\oleg}[1]{{\it [Oleg says: #1]}}

@@

\section{CPS and the problem of generating names}
\omitnow{
In this paper, we are concerned with building and composing
generators. We start with the code expressing the algorithm (in
OCaml), and, adding MetaOCaml annotations, turn that code into a
\emph{code expression} that will generate the code when runs. We then
modify that code expression to add various parameters and aspects, to
yield the complete generator.

@@

To solve the problem of writing a generator for a |let| expression 
when the generated body is not yet known, we use CPS.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% And now, a better -- and shorter -- write-up.

@@

The |do|-notation of Haskell (cite report?) shows that it is possible
to write essentially CPS code in the conventional-looking style. The
|do| notation is the notation for writing monadic code (cite Moggi?).
The benefit of monadic notation is not that it can represent CPS (cite
Filinski), but it helps in composability by offering to add different
layers of effects (state, exception, non-determinism, etc) to the
basic monad (cite Hudak, Peyton-Jones).

A monad (cite Moggi?) is an abstract datatype representing
computations that yield a value and may have an \emph{effect}.
The datatype must have at least two operations, |return| to build
trivial effect-less computations and |bind| for combining
computations. These operations must satisfy \emph{monadic laws}:
|return| being the left and the right unit of |bind| and |bind| being
associative. Figure XXX defines the monad in the present paper
and shows its implementation.

\begin{figure*}
\begin{code}
type ('v,'s,'w) monad = 's -> ('s -> 'v -> 'w) -> 'w
let ret a = fun s k -> k s a
let bind a f = fun s k -> a s (fun s' b -> f b s' k)
let fetch s k = k s s
let store v s k = k (v::s) ()

let k0 s v = v
let runM m = m [] k0

let retN a = fun s k -> .<let t = .~a in .~(k s .<t>.)>.
\end{code}
\caption{Our monad}
\end{figure*}

Our monad represents two kinds of computational effects: reading and
writing a computation-wide state, and control effects. The latter are
normally associated with exceptions, forking of computations, etc. --
in general, whenever a computation ends with something other than
invoking its natural continuation in the tail position. In our case
however the control effects manifest themselves as code generation.

In Figure XXX, the monad is implemented as a function of two
arguments: the state (of type |s|) and the continuation. The
continuation receives the current state, the value (of the type |v|) and
yields the answer of the type |w|.  The monad is polymorphic over all
these three type parameters. Other implementations are
possible. Except the code in Figure XXX, the rest of our code treats
the monad as a truly abstract data type. The implementation of basic
monadic operations |ret| and |bind| are conventional. It is easy to
see that the monadic laws are satisfied. Other monadic operations
construct computations that do have effects. Operations |fetch| and
|store v| construct computations that read and write the state. In our
case the state is a list (of polymorphic variants), which models an
open discriminated union, as we shall see later.

@@

Finally, |runM| runs our monad, that is, performs the computation of
the monad and get the result, which in our case is the code
expression. We run the monad by passing it the initial state and the
initial continuation |k0|. We can now re-write our |param_code2|
example of the previous section as

\begin{code}
let param_code3 plus one =
  let gen x y = bind (retN (plus y one)) (fun ce -> 
                ret (plus ce (plus x ce)))
  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
\end{code}
% param_code3 plus one;;

That does not seem like the improvement. But here we can introduce the
do notation (Appendix A), which is patterned after the do-notation of Haskell.

\begin{code}
let param_code4 plus one =
  let gen x y = mdo { ce <-- retN (plus y one);
                      ret (plus ce (plus x ce)) }
  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
\end{code}

@@

Note that the monad is similar to the one used in \cite{KiselyovTaha}.
However, the latter work used only |retN| of all monadic operations,
and used fixpoints (for performing iterations at the generation time).
In this paper we do not use monadic fixpoints (because the generator
is not recursive) but we make extensive use of monadic operations for
generating conditional and looping operations.

@@

\section{Appendix A}
The do notation for Ocaml.
We use camlp4. Show the grammar. Explain the difficulty: although
nice, the do notation cannot actually be expressed in a LR(n) grammar.
Perhaps give the command line how to load the camlp4 module into
ocaml. Point to the code.
\end{document}
